### 2016-07-21 Projects/DelCo_2012/Report/DelCo_2012.tex
Completed discussion of gas cooling
working on discussion of fire suppression

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$ 15~\%.  In other words, for a difference between one measurement and another measurement to be considered significantly different, the difference would have to be greater by more than $\pm$ 15~\%.  If the difference between two measurements was $\pm$ 15~\% or less, the results are considered to be similar.

As pointed out in the results section, the conditions at TC array 3 represented the "best" gas cooling results with the least impact of agents hitting and cooling the thermocouples themselves.

Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAFS stream had no discernible impact, either positive or negative, in the gas cooling experiments.
At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.

@@

\subsection{Iowa State University Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate in gpm is equal to the volume of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
required fire flowrate = Volume / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the  based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). Our fire rooms ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore the required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 168~Lpm (44~gpm).

\subsection{National Fire Academy (1918)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, in gpm, equal to the floor area (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
required fire flowrate = Area / 3
\end{equation}

The fire rooms in these experiments ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$). Applying the NFA formula based on floor area yields an required flowrate for the single story fire room of 369~Lpm (97~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm).

>>>>

The estimated total expanded uncertainty of the temperature measurements in these experiments is $\pm$ 15~\%.  In other words, for a difference between one measurement and another measurement to be considered significantly different, the difference would have to be greater by more than $\pm$ 15~\%.  If the difference between two measurements was $\pm$ 15~\% or less, the results are considered to be similar. That said, the estimated uncertainty is based on previous studies where suppression with a water based agent was included.

As pointed out in the results section, the conditions at TC Array 3 represented the ``'best'' gas cooling results with the least impact of agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAFS stream had no discernible impact, either positive or negative, in the gas cooling experiments.  At the same time there was no measurable negative effect of the air bubbles with regard to gas cooling.

The data from the thermocouples in TC Array 2, provide a sense of the impact of a combination of surface cooling and gas cooling.  For example, when the nozzle was in the "middle position" the straight or solid hose streams would impact near the middle of the ceiling of the room and they would create a fan pattern across the ceiling that would cause the majority of the water to impact the location of TC array 2 and the adjacent walls.  In this location, the data tended to favor water.  But again this did not represent gas cooling, this represented water or CAF hitting the thermocouples and cooling the surrounding gases. As a result the temperature reductions of the thermocouples in TC Array 2 are higher than those of the thermocouples in TC Array 3, as shown in \ref{fig:CAFS_Water_A2_mid}, and \ref{fig:CAFS_Water_A3_mid}.

TC Array 1 had the worse data fit or the widest spread of temperature change because sometimes, some of the thermcouples in the array were getting impacted by water or CAF while during other nozzle streams they were not. Examining the spray density figures in Appendix A.1, provides a range water collected in the bucket where TC Array 1 was located.  In many cases such as CAF Straight Stream Mid, the foam solution collected was near zero. In other cases such as CAF Fog Mid, the amount of foam solution colected was more than 3~kgs.  Similar variations occurred with water.  As a result, TC Array 1 did not provide consistant results for gas cooling.

The data from this test series does not support the premise of firefighters being injured during an interior firefighting operation with CAF, because it was thought to provide less cooling than a similar hose stream of plain water.

@@

\subsection{Iowa State University Fire Flow Formula (1963)}
The Iowa Formula for determining the required fire flowrate for a given structure was developed, by Keith Royer and Bill Nelson, based on theoretical heat absorption capabilities of water. The required fire flowrate in gpm is equal to the volume of the fire compartment (in cubic feet) divided by 100:
\begin{equation}\label{eq:isu_form}
required fire flowrate = Volume / 100
\end{equation}
Due to inefficiencies in applying the water, some feel that the fire flow value from the Iowa State University formula should be multiplied by a factor of two to four~\cite{NFPA}. Applying the  based on floor area yields an application density of 3.3~Lpm/m$^2$ (0.08~gpm/ft$^2$) for ceiling heights of 2.44~m (8~ft) or 4.1~Lpm/m$^2$ (0.10~gpm/ft$^2$) for ceiling heights of 3.05~m (10~ft). Our fire rooms ranged in floor area from approximately 66~m$^3$ (2330~ft$^3$) to 125~m$^3$ (4410~ft$^3$). Therefore the required flowrate for the fire room in the single story structure based on the Iowa State University formula would be approximately 87~Lpm (23~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 168~Lpm (44~gpm).  While the formula has been widely taught over the years, there is some uncertainty about the application of the formula.  In some texts it mentions that the Iowa Formula provides the amount of water that needs to be applied in 30 seconds.  If that is the case then the flowrates given "per minute" must be doubled.  So in our case the required flowrates would increase to 174~Lpm (46~gpm)
and 336~Lpm (88 gpm), respectively.  During the experiments the flowrate was approximately, 454~Lpm (120~gpm), which exceeds the required flowrate of the Iowa Formula.  So the fact that all of the fires were quickly controlled is consistant with this methodology.

\subsection{National Fire Academy (1981)}
The National Fire Academy (NFA) developed a formula based on a study of fire flows, which were successful in controlling a large number of working fires~\cite{Klaene:1}. This formula sets the required fire flow, in gpm, equal to the floor area (in square feet) divided by 3.
\begin{equation}\label{eq:nfa_form}
required fire flowrate = Area / 3
\end{equation}

The fire rooms in these experiments ranged in floor area from approximately 27~$m^2$ (291~$ft^2$) to 51.4~$m^2$ (553~$ft^2$). Applying the NFA formula based on floor area yields an required flowrate for the single story fire room of 369~Lpm (97~gpm). The required flowrate for the fire room in the lower level of the two story would be approximately 700~Lpm (184~gpm).  Comparing our hosestream flowrates to the NFA Formula requirements indicates that 454~Lpm (120~gpm) was sufficient for the smaller room fire in the single story but was below the reuired flowrate of approximately 700~Lpm (184~gpm) for the large fire compartment in the two story structure.  This is not consistant with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~Lpm (120~gpm).

Due to recent research conducted by NIST and the UL Firefighter Safety Research Institute (FSRI), the fire service has a re-newed interest in controlling the air available to a ventilation limited compartment fire.  By limiting the air (oxygen) available for combustion is one means of limting the heat release rate.  If the heat release rate of a fire is reduced due to a lack of oxygen needed for combustion, then in theory the amount of water needed to cool and suppress the fire should also be reduced.

Both of the formulas listed above do not overtly account for the ventilation of the fire structure only the size of the fire structure.  Further the most recent formula was developed in 1981, based on the fire ground experience of fire chiefs.  The experience that was shared was likely from fire that occurred in the 1960's and the 1970's.  It has been demonstrated that the synthetic fuels which fill the homes of the 2000's burn faster and hotter (higher heat release rate) than the natural fuels that filled the homes in the 1940's and 1950's.  The change of materials in our homes began in the 1960's and continues through to present day.  However it was unlikely that this change was factored in to the NFA Formula.

All of the fire conditions generated during these experiments were ventilation limited or ventilation controlled fires. In other words if more oxygen was supplied to the fuel gases available in the fire compartment and more products of combustion were exhausted from the fire compartment, the heat release rate of the fire would have increased. There are simple formulas which may provide some insight here, although for full disclosure these simple algorithms have not been validated with the geometry that we used in these experiments.

If we have a single open doorway to a residential scale room, which is approximately 0.9~m (3 ft) wide and 2.1~m (6.9 ft) high, the minimum heat release rate needed to flashover the room would be about 2 MW and the amount of heat release rate that the doorway could support assuming complete combustion of the oxygen that would flow through the doorway opening via natural convection would be approximately 4 MW.  If we double the width of the open doorway (the doorway to the lower level of the two story structure)  Then the heat release rate that could be generated inside the compartment, this does not account for flames burning outside of the compartment, would be approximately 8 MW.
